Towards Compiler-Agnostic Performance in Finite-Difference Codes

### 2015-09-15 docs/papers/shallow_performance/PSEA-2015_review_response.txt

> 5) p3: What does "retaining the I/O capability" imply? Is bitwise
> reproducability of results a requirement?

It simply implies that the solution fields can be written to disk at a
user-defined interval. Text has been added to this effect. Bit-wise
reproducibility is not a requirement.

> 6) p4: it is unclear from Figure 1 and the description on the previous
> page whether invoke_tstep is a routine written by the scientist and
> 'transformed' or entirely the responsibility of PSyKAI. It the latter
> is the case, how can PSyKAI determine which kernels to call, in which
> order, which parameters to pass etc.?

PSyKAl is a design approach, not a tool. Figure 1 illustrates the
structure of the Shallow code after it has been re-structured
following PSyKAl principles. How the code in the different layers
is produced is not the subject of this paper.

@@

These recipes are implemented as python scripts that apply PSyclone
Transformations.

General comments / suggestions
------------------------------

> 1) There seems to be no preprint available yet for reference [2]

RUPERT/GRAHAM

> 2) Figure 2/3: A better visual grouping of the bars by platform and
> compiler e.g. by using colour to differentiate compilers and hatching
> to different platforms would help with interpretation of the plots
> 3) Figure 4: The bars per cluster are quite hard to tell apart visually

We would love to use colour but the formatting guidlines say we
can't. However, we have added additional labelling to what was Figure
3 in order to aid interpretation.

### 2015-10-01 docs/papers/shallow_performance/shallow_perf.tex
Modifications to the paper to provide a little information about how PSyKAl relates to OP2 and to fix various typos and inconsistencies.

\begin{abstract}
In this paper we evaluate the performance implications of applying a
technique which we call \psykal\ to finite difference Ocean models. 
%
In \psykal, the code related to the underlying science is formally
separated from code related to parallelisation and single core
optimisations. This separation of concerns allows scientists to code
their science independently of the underlying hardware architecture
(thereby keeping a single code base) and for optimisation specialists
to be able to tailor the code for a particular machine independently
of the science code.
%
A finite difference shallow water benchmark optimised for cache based
architectures is taken as the starting point. A vanilla \psykal\ version
is written and the performance of the two compared. The optimisations
that were applied to the original benchmark (loop fusion {\it etc.}) are then
applied to the \psykal\ version as a set of code modifications to the
optimisation layer. Performance results are presented for the Cray,
Intel and GNU compilers on Intel Ivybridge and Haswell processors and
for the IBM compiler on Power8.
%
Results show that the combined set of code modifications obtain
performance that is within a few percent of the original code for all
compiler and architecture combinations on all tested problem
sizes. The only exception to this (other than where we see significant
speed-ups) is the Gnu compiler on Haswell for one problem size. Our
tests indicate that this may be due to immature support for that
architecture in the Gnu compiler -- no such problem is seen on the Ivy
Bridge system.
%%
Therefore, the \psykal\ approach can be used with negligable
performance loss and sometimes small performance gains compared to the
original optimised code, there is no single best hand optimised
implementation of the code for all the compilers tested, and the
flexibility in applying code modifications to the optimisation layer
in the \psykal\ version allows for a significant performance improvement
on the IBM Power8 compared with the original code.

@@

While this approach is general, we are currently applying it to
Atmosphere and Ocean models written in Fortran where domain
decomposition is typically performed in the latitude-longitude
direction, leaving columns of elements on each domain-decomposed
partition.

The top layer, in terms of calling hierarchy, is the Algorithm
layer. This layer specifies the algorithm that the scientist would like
to perform (in terms of calls to kernel and infrastructure routines)
and logically operates on full fields. We say logically here as the
fields may be domain decomposed, however the algorithm layer is not
aware of this. It is the scientist's responsibility to write this
algorithm layer.

@@

As the PSy layer iterates over columns, the single core performance
can be optimised by applying transformations such as loop fusion and
loop blocking. Additionally, the potential parallelism within this
iteration space can also be optimised and parallelised. The PSy layer
can therefore be tailored for a particular hardware (such as
multi-core, many-core, GPGPUs, or some combination thereof) and
software (such as compiler, operating system, MPI implementation)
configuration with no change to the algorithm or kernel layer
code. This approach therefore offers the potential for portable
performance.

@@

For this work we use a benchmark called Shallow which solves the
shallow-water equations on a bi-periodic plane following the
finite-difference scheme introduced by Sadourny~\cite{sadourny75}.
This software was originally written in 1984 by Paul Swarztrauber of
the National Center for Atmospheric Research, US.  However, in common
with many scientific codes, it has subsequently undergone some sort of
evolutionary development with subsequent people making various changes
and optimising it for previous generations of hardware.  There is no
complete record of the people involved and the precise hardware they
were targetting. In describing our work, we term the version of the
Shallow program obtained at the beginning of this project the
`original' version.

@@

In order to maximise the flexibility (and thus potential for
optimisation) of the \psykal\ version of Shallow, we made the kernels
as fine-grained as possible. In this case, this resulted in eight
distinct kernels, each of which operated on a single field at a single
point (since we have chosen to use point-wise kernels). With a little
bit of tidying/re-structuring, we found it was possible to express the
contents of the main time-stepping loop as a single invoke (a call to
the PSy layer) and a call to the I/O system
(Figure~\ref{FIG_psykal_shallow_structure}). The single PSy-layer
routine then consists of applying each of the kernels to all of the
points on the model mesh requiring an update. In its basic,
unoptimised (`vanilla') form, this PSy-layer routine then contains a
doubly-nested loop around each kernel call, as indicated by the
pseudo-code in Figure~\ref{FIG_psykal_shallow_structure}.

@@

As with any full oceanographic model, boundary conditions must be
applied at the edges of the model domain. In the case of Shallow we
have periodic boundary conditions in both the $x$- and $y$-directions.
These are simply implemented by having additional (`halo')
rows/columns at the edges of the domain and ensuring the data in them
is up-to-date before it is read. These rows/columns are updated by
copying data from the corresponding row/column on the opposite side of
the domain. In the current work, the lines of code to do these copies
are manually inserted in the PSy-layer routine as required.

@@

\begin{table}[!t]
% increase table row spacing, adjust to taste
\renewcommand{\arraystretch}{1.3}
\caption{The matrix of compilers and CPUs used in this work. The use
  of a compiler on a given CPU is indicated by the specification of
  the version of the compiler in the relevent element. No entry
  indicates that a compiler was not available/used on that CPU.}
\label{TABLE_compilers}
\centering
\begin{tabular}{|l|c|c|c|c|}
\hline
                 & \multicolumn{4}{c|}{Compiler}             \\
\hline
                 & Gnu   & Intel       & Cray    & IBM     \\
\hline
Intel Haswell    & 4.9.3 & 14.0.0      &         &          \\
Intel Ivy Bridge & 4.9.1 & 14.0.1.106  & 8.3.3   &          \\
IBM Power 8      &       &             &         & 15.1.2     \\
\hline
\end{tabular}
\end{table}

@@

Beginning with the vanilla \psykal\ version, we then manually apply a
series of (cumulative) code transformations while obeying the \psykal\ 
separation of concerns, {\it i.e.} optimisation is restricted to the
middle, {PS}y layer and leaves the kernel and algorithm layers
unchanged. The aim of these optimisations is to recover, as much as is
possible, the structure and thus performance of the original version
of the code. We began with two transformations that affect all of the
kernels. After this, we determined the transformations to perform by
comparing the timing profiles of the \psykal\ and original versions of
the code. We then tackled the sections resposible for the largest
slowdowns and iterated this process. While doing this we also made
heavy use of the optimisation reports produced by the compilers and
how these changed when moving from the original to the \psykal\ version
of the code. The transformations we have performed are as follows:

@@

\item Manually in-line all kernel bodies into the middle-layer
  subroutine (kernel inlining);

@@

We now examine the performance of the \psykal\ version of Shallow and
the code transformations that are required to reach the levels
obtained by the original.  We do this for the $256^{2}$ problem size
since this fits within LLC on all of the CPUs we are using here.  The
performance of the \psykal\ version after each transformation has been
manually applied to the code is plotted in
Figure~\ref{FIG_opt_stages_256}. Results for a given compiler/CPU are
given relative to the performance of the original version of Shallow
with the same compiler/CPU.

@@

With the Gnu compiler (first and third clusters in
Figure~\ref{FIG_opt_stages_256}), the vanilla \psykal\ version of
Shallow only achieves ~5-6\% of the performance of the
original. Simply copying the code for each of the kernel subroutines
into the module from which they are called (module in-lining, T3) has the
most dramatic effect on the performance of the compiled code; it now
achieves more than 80\% of the performance of the original. The
transformation(s) required to get more than 90\% then become
architecture dependent: on Haswell it is just loop fusion (T4) while on Ivy
Bridge we do not exceed 90\% until we have done all of the steps up to
and including T8; fully fusing (fusing the inner loops of)
the first loop nest. The dramatic performance improvement obtained
from the module-inlining (T3) demonstrates that the Gnu
compiler is unable to optimise over separate source files.

@@

The fifth cluster in Figure~\ref{FIG_opt_stages_256} shows the
evolution of the performance of the \psykal\ version of Shallow with
the Cray compiler. Again, this has significant differences from the
behaviour seen with the Gnu and Intel compilers. As with the Intel
compiler, the performance of the Vanilla \psykal\ version, T1,  is fairly
good at 74\% of that of the original. In contrast to the other
compilers consisdered so far, the first significant transformation is
simply to specify the bounds of the arrays being used within the
middle layer using module variables (as opposed to specifying them as
assumed size). This tells the compiler that all of the arrays used in
the middle layer have the same extent and also that all of the
computational loops are over almost all elements of these arrays. This
simple change gives a \psykal\ version that achieves 88\% of the
performance of the original.

Subsequent transformations actually harm performance until T6 where
the field-copy operation (lower loop in
Figure~\ref{FIG_time_smooth_code}) is fused with the preceeding
loop. The resulting \psykal\ version now achieves 94\% of the original
performance. Two further steps are required to match the latter: the
first loop nest is fully fused (T8 - both inner and outer loops fused)
and second, the array/loop bounds are passed as an argument to the
middle layer (T9). This is done in order to indicate to the compiler
that these bounds remain constant for the duration of each iteration
of the time-stepping loop.

@@

Finally, we compare the performance of the best \psykal\ version of
Shallow with that of the original. In
Figure~\ref{FIG_slowdown_summary} we plot the percentage
  difference between the performance of the original and the (best)
\psykal\ versions of Shallow for each compiler/CPU combination. The
most important feature of this plot is that the performance penalty
incurred by the \psykal\ version is less than four percent with only
one exception where it is 6.6\%.  In fact, for the Intel and Cray
compilers, the \psykal\ version of Shallow is never more than 2\%
slower than the original and in some cases is faster.  This
demonstrates that we can reliably recover the performance of the
original version of the code, despite the significant restructuring
required by the \psykal\ approach.

@@

Turning to the cases where the \psykal\ version is significantly
{\em faster} than the original, we see from
Figure~\ref{FIG_slowdown_summary} that, with the exception of Power 8,
this only occurs for the smallest domain size. This is no doubt due in
part to the fact that the problem executes very quickly and therefore
a small reduction in run-time will show as a large precentage. Since
this problem is well contained in L2 cache, its performance will also
be more sensitive to small differences, {\it e.g.} in the way that CPU
registers are used.

The performance of the Power 8 system with the $512^2$ domain size is
somewhat anomolous: the \psykal\ version of the code performs
significantly better than the original and in addition, this problem
size is also the only one where the Power 8 out-performs the Haswell
system (see Figure~\ref{FIG_orig_perf_summary}) for the original
version.  The results in Figure~\ref{FIG_orig_perf_summary} indicate
that the $512^2$ case marks a transition from the working set being in
cache (sizes $\leq 256^2$) to spilling out of cache (size $1024^2$). We
therefore suggest that cache is being used more efficiently on the
Power 8 system and this is further enhanced by the IBM-supplied
transformations in the \psykal\ version.

@@

In particular, the results on the IBM system demonstrate two key points.
%
First, even for a relatively simple
benchmark code, the code structure required to obtain good performance
between different architecture and/or compiler combinations may
differ. For hand optimised codes this implies the need to support
multiple code bases in order to achieve portable performance.
%
Second, the \psykal\ approach directly addresses the problem of
requiring different optimisations for different architectures and/or
compilers by limiting these differences to a separate performance
layer, thereby allowing the natural science code base to remain
unchanged.

In future work we will extend our performance portability analysis of
the \psykal\ approach to shared memory parallelisation optimisations on
different architectures using different parallelisation techniques (in
particular OpenMP on multi/many-core and OpenACC on GPU's).
%
We will then analyse the performance of a domain-specific compiler
that is currently under development. This compiler will generate
optimised PSy-layer code using a user-provided recipe of
transformations, thereby removing the need for optimisation experts to
manually write the PSy layer.

% use section* for acknowledgement
\section*{Acknowledgments}

>>>>

\begin{abstract}
In this paper we evaluate the performance implications of applying a
technique which we call \psykal\ to finite difference Ocean models. 
%
In \psykal, the code related to the underlying science is formally
separated from code related to parallelisation and single core
optimisations. This separation of concerns allows scientists to code
their science independently of the underlying hardware architecture
(thereby keeping a single code base) and for optimisation specialists
to be able to tailor the code for a particular machine independently
of the science code.
%
A finite difference shallow water benchmark optimised for cache based
architectures is taken as the starting point. A vanilla \psykal\ version
is written and the performance of the two compared. The optimisations
that were applied to the original benchmark (loop fusion {\it etc.}) are then
applied to the \psykal\ version as a set of code modifications to the
optimisation layer. Performance results are presented for the Cray,
Intel and GNU compilers on Intel Ivybridge and Haswell processors and
for the IBM compiler on Power8.
%
Results show that the combined set of code modifications obtain
performance that is within a few percent of the original code for all
compiler and architecture combinations on all tested problem
sizes. The only exception to this (other than where we see performance
improvement) is the Gnu compiler on Haswell for one problem size. Our
tests indicate that this may be due to immature support for that
architecture in the Gnu compiler -- no such problem is seen on the Ivy
Bridge system. Further, the original code performed poorly using the
IBM compiler on Power8 and needed to be modified to obtain performant
code.
%%
Therefore, the \psykal\ approach can be used with negligible
performance loss and sometimes small performance gains compared to the
original optimised code, there is no single best hand optimised
implementation of the code for all the compilers tested, and the
flexibility in applying code modifications to the optimisation layer
in the \psykal\ version allows for a significant performance improvement
on the IBM Power8 compared with the original code.

@@

In common with OP2, the \psykal approach separates out the science
code and the performance related code and does this via a separate
layer. The calls that specify parallelism in both approaches are
similar in terms of where they are placed in the code and in their
semantics but the \psykal approach supports the specification of more
than one Kernel in a parallel region of code, compared with one for
OP2, giving more scope for optimisation. Another difference is that
the description metadata is provided with the Kernel in the \psykal
approach whereas it is provided as part of the Kernel call in OP2.

While the \psykal approach is general, we are currently applying it to
Atmosphere and Ocean models written in Fortran where domain
decomposition is typically performed in the latitude-longitude
direction, leaving columns of elements on each domain-decomposed
partition.

The top layer, in terms of calling hierarchy, is the Algorithm
layer. This layer specifies the algorithm that the scientist would like
to perform (in terms of calls to kernel and infrastructure routines)
and logically operates on full fields. We say logically here as the
fields may be domain decomposed, however the Algorithm layer is not
aware of this. It is the scientist's responsibility to write this
Algorithm layer.

@@

As the PSy layer iterates over columns, the single core performance
can be optimised by applying transformations such as loop fusion and
loop blocking. Additionally, the potential parallelism within this
iteration space can also be optimised and parallelised. The PSy layer
can therefore be tailored for a particular hardware (such as
multi-core, many-core, GPGPUs, or some combination thereof) and
software (such as compiler, operating system, MPI implementation)
configuration with no change to the Algorithm or Kernel layer
code. This approach therefore offers the potential for portable
performance.

@@

For this work we use a benchmark called Shallow which solves the
shallow-water equations on a bi-periodic plane following the
finite-difference scheme introduced by Sadourny~\cite{sadourny75}.
This software was originally written in 1984 by Paul Swarztrauber of
the National Center for Atmospheric Research, US.  However, in common
with many scientific codes, it has subsequently undergone some sort of
evolutionary development with subsequent people making various changes
and optimising it for previous generations of hardware.  There is no
complete record of the people involved and the precise hardware they
were targeting. In describing our work, we term the version of the
Shallow program obtained at the beginning of this project the
`original' version.

@@

In order to maximise the flexibility (and thus potential for
optimisation) of the \psykal\ version of Shallow, we made the kernels
as fine-grained as possible. In this case, this resulted in eight
distinct kernels, each of which operated on a single field at a single
point (since we have chosen to use point-wise kernels). With a little
bit of tidying/re-structuring, we found it was possible to express the
contents of the main time-stepping loop as a single invoke (a call to
the PSy layer) and a call to the I/O system
(Figure~\ref{FIG_psykal_shallow_structure}). The single PSy layer
routine then consists of applying each of the kernels to all of the
points on the model mesh requiring an update. In its basic,
unoptimised (`vanilla') form, this PSy layer routine then contains a
doubly-nested loop around each kernel call, as indicated by the
pseudo-code in Figure~\ref{FIG_psykal_shallow_structure}.

@@

As with any full oceanographic model, boundary conditions must be
applied at the edges of the model domain. In the case of Shallow we
have periodic boundary conditions in both the $x$- and $y$-directions.
These are simply implemented by having additional (`halo')
rows/columns at the edges of the domain and ensuring the data in them
is up-to-date before it is read. These rows/columns are updated by
copying data from the corresponding row/column on the opposite side of
the domain. In the current work, the lines of code to do these copies
are manually inserted in the PSy layer routine as required.

@@

\begin{table}[!t]
% increase table row spacing, adjust to taste
\renewcommand{\arraystretch}{1.3}
\caption{The matrix of compilers and CPUs used in this work. The use
  of a compiler on a given CPU is indicated by the specification of
  the version of the compiler in the relevant element. No entry
  indicates that a compiler was not available/used on that CPU.}
\label{TABLE_compilers}
\centering
\begin{tabular}{|l|c|c|c|c|}
\hline
                 & \multicolumn{4}{c|}{Compiler}             \\
\hline
                 & Gnu   & Intel       & Cray    & IBM     \\
\hline
Intel Haswell    & 4.9.3 & 14.0.0      &         &          \\
Intel Ivy Bridge & 4.9.1 & 14.0.1.106  & 8.3.3   &          \\
IBM Power 8      &       &             &         & 15.1.2     \\
\hline
\end{tabular}
\end{table}

@@

Beginning with the vanilla \psykal\ version, we then manually apply a
series of (cumulative) code transformations while obeying the \psykal\ 
separation of concerns, {\it i.e.} optimisation is restricted to the
middle, {PS}y layer and leaves the Kernel and Algorithm layers
unchanged. The aim of these optimisations is to recover, as much as is
possible, the structure and thus performance of the original version
of the code. We began with two transformations that affect all of the
kernels. After this, we determined the transformations to perform by
comparing the timing profiles of the \psykal\ and original versions of
the code. We then tackled the sections responsible for the largest
slowdowns and iterated this process. While doing this we also made
heavy use of the optimisation reports produced by the compilers and
how these changed when moving from the original to the \psykal\ version
of the code. The transformations we have performed are as follows:

@@

\item Manually in-line all kernel bodies into the middle layer
  subroutine (kernel inlining);

@@

We now examine the performance of the \psykal\ version of Shallow and
the code transformations that are required to reach the levels
obtained by the original.  We do this for the $256^{2}$ problem size
since this fits within LLC on all of the CPUs we are using here.  The
performance of the \psykal\ version after each transformation has been
manually applied to the code is plotted in
Figure~\ref{FIG_opt_stages_256}. Results for a given compiler/CPU are
presented relative to the performance of the original version of Shallow
with the same compiler/CPU.

@@

With the Gnu compiler (first and third clusters in
Figure~\ref{FIG_opt_stages_256}), the vanilla \psykal\ version of
Shallow only achieves ~5-6\% of the performance of the
original. Simply copying the code for each of the kernel subroutines
into the module from which they are called (module in-lining, T3) has the
most dramatic effect on the performance of the compiled code; it now
achieves more than 80\% of the performance of the original. The
transformation(s) required to get more than 90\% then become
architecture dependent: on Haswell it is just loop fusion (T4) while on Ivy
Bridge we do not exceed 90\% until we have applied all of the steps up to
and including T8; fully fusing (fusing the inner loops of)
the first loop nest. The dramatic performance improvement obtained
from the module-inlining (T3) demonstrates that the Gnu
compiler is unable to optimise over separate source files.

@@

The fifth cluster in Figure~\ref{FIG_opt_stages_256} shows the
evolution of the performance of the \psykal\ version of Shallow with
the Cray compiler. Again, this has significant differences from the
behaviour seen with the Gnu and Intel compilers. As with the Intel
compiler, the performance of the Vanilla \psykal\ version, T1,  is fairly
good at 74\% of that of the original. In contrast to the other
compilers considered so far, the first significant transformation is
simply to specify the bounds of the arrays being used within the
middle layer using module variables (as opposed to specifying them as
assumed size). This tells the compiler that all of the arrays used in
the middle layer have the same extent and also that all of the
computational loops are over almost all elements of these arrays. This
simple change gives a \psykal\ version that achieves 88\% of the
performance of the original.

Subsequent transformations actually harm performance until T6 where
the field-copy operation (lower loop in
Figure~\ref{FIG_time_smooth_code}) is fused with the preceding
loop. The resulting \psykal\ version now achieves 94\% of the original
performance. Two further steps are required to match the latter: the
first loop nest is fully fused (T8 - both inner and outer loops fused)
and second, the array/loop bounds are passed as an argument to the
middle layer (T9). This is done in order to indicate to the compiler
that these bounds remain constant for the duration of each iteration
of the time-stepping loop.

@@

Finally, we compare the performance of the best \psykal\ version of
Shallow with that of the original for different problem sizes. In
Figure~\ref{FIG_slowdown_summary} we plot the percentage
  difference between the performance of the original and the (best)
\psykal\ versions of Shallow for each compiler/CPU combination. The
most important feature of this plot is that the performance penalty
incurred by the \psykal\ version is less than four percent with only
one exception where it is 6.6\%.  In fact, for the Intel and Cray
compilers, the \psykal\ version of Shallow is never more than 2\%
slower than the original and in some cases is faster.  This
demonstrates that we can reliably recover the performance of the
original version of the code, despite the significant restructuring
required by the \psykal\ approach.

@@

Turning to the cases where the \psykal\ version is significantly
{\em faster} than the original, we see from
Figure~\ref{FIG_slowdown_summary} that, with the exception of Power 8,
this only occurs for the smallest domain size. This is no doubt due in
part to the fact that the problem executes very quickly and therefore
a small reduction in run-time will show as a large percentage. Since
this problem is well contained in L2 cache, its performance will also
be more sensitive to small differences, {\it e.g.} in the way that CPU
registers are used.

The performance of the Power 8 system with the $512^2$ domain size is
somewhat anomalous: the \psykal\ version of the code performs
significantly better than the original and in addition, this problem
size is also the only one where the Power 8 out-performs the Haswell
system (see Figure~\ref{FIG_orig_perf_summary}) for the original
version.  The results in Figure~\ref{FIG_orig_perf_summary} indicate
that the $512^2$ case marks a transition from the working set being in
cache (sizes $\leq 256^2$) to spilling out of cache (size $1024^2$). We
therefore suggest that cache is being used more efficiently on the
Power 8 system and this is further enhanced by the IBM-supplied
transformations in the \psykal\ version.

@@

More unpredictably, the code structure required to obtain good
performance for different architecture and/or compiler combinations
differs, even for this relatively simple benchmark code. For hand
optimised codes this implies the need to support multiple code bases
in order to achieve portable performance.
%
The \psykal\ approach directly addresses the problem of requiring
different optimisations for different architectures and/or compilers
by limiting these differences to a separate performance layer, thereby
allowing the natural science code base to remain unchanged.

In future work we will extend our performance portability analysis of
the \psykal\ approach to shared memory parallelisation optimisations on
different architectures using different parallelisation techniques (in
particular OpenMP on multi/many-core and OpenACC on GPU's).
%
We will then analyse the performance of a domain-specific compiler
that is currently under development. This compiler will generate
optimised PSy layer code using a user-provided recipe of
transformations, thereby removing the need for optimisation experts to
manually write the PSy layer.

% use section* for acknowledgement
\section*{Acknowledgements}
